# Chapter 6: Analysis of longitudinal data

- It would have been nice to be able to get more out of this weeks exercises and materials; it is easy to simply copy and modify the code, but I would like to learn much more

```{r message=FALSE, warning=FALSE}
date()

# packages in use
library(dplyr)
library(ggplot2)
library(tidyr)

```

## 6.1 RATS

In this subchapter I will implement the analyses of Chapter 8 of MABS using the RATS data.

### 6.1.1 Data

The RATS data utilized here is drawn from a nutrition study conducted with rats ([see Crowder & Hand 1990](https://www.routledge.com/Analysis-of-Repeated-Measures/Crowder-Hand/p/book/9780367450847)). In the experiment the rats were grouped into three groups which had different diets, and each rats’s weight (grams) was recorded repeatedly over a 9-week
period (about weekly, with an exception of week seven when two measurement were taken). 

The original RATS data included 16 observations of 13 variables, that is 16 rats with multiple measurements over time (and a few other variables). Here, the data is transformed from wide to long format by time (measurements), meaning that each time point (i.e. measurement) is forms an observation and, thus, every individual rat has as many observations as there are measurements. Thererore, in the long format there are 176 observations of 4 variables, which are:

- ID: unique id of participants (rats)
- Group: the treatment groups 1,2 and 3
- Weight: rats' weight in grams
- Time: time point (days) of the weight measurement

NB: The variable 'WD' is removed from the data set as its information is already included in the 'Time' variable and thus there is no use for it. In addition, ID and Group variables are transformed to factors.

```{r message=FALSE, warning=FALSE}
# load the data in long format
rats_long <- read.table("data/rats_long.csv", sep = ";")

# prepare data for the analysis
## remove WD variable 
## Group and ID -> factors
rats_long <- rats_long %>%
  select(-WD) %>%
  mutate(Group = as.factor(Group),
         ID = as.factor(ID))

str(rats_long)
head(rats_long)
glimpse(rats_long)
```

### 6.1.2 Graphical examination of the group based differences

First, I draw a plot for every observed rat; the plot shows measured weight (y-axis) over time (x-axis) for every rat (lines). Rats in every group have gained weight during the experiment, although the growth seems to be bigger in groups 2 and 3 compared to group 1. However, the plot also illustrates that in the beginning of the experiment (time 0) the rats differ from each other in terms of the weight and bigger rats are bigger throughout the experiment. Rats in group 1 are much smaller than those in the groups 2 and 3. Also there is one much bigger individual in the group 2. 

```{r message=FALSE, warning=FALSE}

# Draw a plot for rats in each group over time
ggplot(rats_long, aes(x = Time, y = Weight, linetype = ID, col = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, times=4)) + #what this does?
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(rats_long$Weight), max(rats_long$Weight))) +
  ggtitle("Individual weight profiles over time by group for the RATS data") +
  xlab("Time (days)")

```

Next, I will standardize/center the weight variable and repeat the above-presented graphical examination. The graphical results are similar, although the trend in weight gain does not seem to be as straightforward as in the previous plot, especially regarding group 3.

```{r message=FALSE, warning=FALSE}

# Standardize the variable weight
rats_long<- rats_long %>%
  group_by(Time) %>%
  mutate(Weight_std = (Weight - mean(Weight)) / sd(Weight)) %>%
  ungroup()

# Draw the plot using the stardardixed weight variable
ggplot(rats_long, aes(x = Time, y = Weight_std, linetype = ID, col = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, times=4)) + #what this does?
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "Standardized weight") +
  ggtitle("Individual weight profiles over time by group for the RATS data after standardization") +
  xlab("Time (days)")


```

To get a better picture of the weight gain and the differences between the groups, I will next produce a graph showing average profiles for each group. Additionally I draw a graph showing side-by-side box plots of the observations at each time point. The graphs illustrate that rats in every groups gain weight over time, although the weight gain is somewhat different in different groups. Figures also shoet that the mean of weights of groups 2 and 3 clearly differs from the group 1. In addition, although there is some overlap between the groups 2 and 3 it seems that they do differ from each other. Moreover, the box plots suggest the presence of multiple possible outliers at a number of time points.

```{r message=FALSE, warning=FALSE}

# Summary data with mean and standard error of weight by treatment group and time

# PROBLEM: DIFFRENT NUMBER OF RATS IN DIFFERENT GROUPS!
## how to take sqrt(n), where n = number of rats in group -> 
### n() which gives the current group size ??
rats_long_s <- rats_long %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight) /  sqrt(n())) %>%
  ungroup()

# Plot the mean profiles
ggplot(rats_long_s , aes(x = Time, y = mean, linetype = Group, shape = Group, col = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.45)) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)") +
  xlab("Time (days)") +
  ggtitle("Mean response profiles for the three groups of rats in the RATS data")

# Alternative mean profile figure: side-by-side box plots of the observations at each time point
ggplot(rats_long, aes(x = as.factor(Time), y = Weight, fill = Group)) +
  geom_boxplot(position = position_dodge(width = 0.9)) +
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position = c(0.9,0.4)) +
  scale_x_discrete(name = "Time (days") +
  ggtitle("Boxplots for the RATS data: weight over time")


```

I will continue the summary approach and look into the post treatment values of the weight. The mean of the all weeks' weight will be the summary measure. Day 1 will be considered here as the baseline and it is excluded from the examination asI guess we are supposed to do. 

```{r}
# Create a summary data by group and subject with mean as the summary variable (ignoring baseline day 1).
rats_summary <- rats_long %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Draw a boxplot of the mean versus group
ggplot(rats_summary, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days 8-64")

```

The box plots presented above shows again that the groups differ from each other and also that every group has an outlier, although only in the case of group two the outlier is quite far from the other means. Thus, I will remove the outlier of group 2 and draw the box plots again. After removing the outlier, the differences between groups are even clearer as can be seen in the picture below. 

```{r message=FALSE, warning=FALSE}
rats_summary_filt <- rats_summary %>%
  filter(mean < 550)

# Draw a boxplot of the mean versus group
ggplot(rats_summary_filt, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days 8-64")
```

### 6.1.3 Statistical examination of the group-based differences

Next, I will first add the baseline measurement (day 1) from the original data to the summary data and then utilize it in fitting a linear model in which the mean is the response variable. When taking the baseline into account results are surprisingly quite different! According to the results from the linear regression model the coefficient of group 2 or 3 are not statistically significant (when considering p<.05 as the threshold). Analysis of variance provides similar results.

**HERE some interpretation**

```{r message=FALSE, warning=FALSE}

## load the data in wide form
rats_wide <- read.table("data/rats_wide.csv", sep = ";")

# Add the baseline (Day 1) from the original data as a new variable to the summary data
rats_summary2 <- rats_summary %>%
  mutate(baseline = rats_wide$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = rats_summary2)

# summary of the fitted model
summary(fit)
# analysis of variance table for the fitted model 
anova(fit)

```



## 6.2 BPRS

*_Implement the analyses of Chapter 9 of MABS using the BPRS data. (0-8 points: 0-4 points for graphs or analysis results + 0-4 points for their interpretations)_*

In this subchapter I will implement the analyses of Chapter 9 of MABS using the BPRS data.

### 6.2.1 Data

```{r message=FALSE, warning=FALSE}
# load the data in long format
bprs_long <- read.table("data/bprs_long.csv", sep = ";")

```

### 6.2.2 Something