# Chapter 6: Analysis of longitudinal data

```{r message=FALSE, warning=FALSE}
date()

# packages in use
library(dplyr)
library(ggplot2)
library(tidyr)

```

## 6.1 RATS

In this subchapter I will implement the analyses of Chapter 8 of MABS using the RATS data.

### 6.1.1 Data

The RATS data utilized here is drawn from a nutrition study conducted with rats ([see Crowder & Hand 1990](https://www.routledge.com/Analysis-of-Repeated-Measures/Crowder-Hand/p/book/9780367450847)). In the experiment the rats were grouped into three groups which had different diets, and each rat's weight (grams) was recorded weekly over a 9-week period (with an exception of week seven when two measurements were taken). 

The original RATS data included 16 observations of 13 variables, that is 16 rats with multiple measurements over time (and a few other variables). Here, the data is transformed from wide to long format by time (measurements), meaning that each time point (i.e. measurement) forms an observation and, thus, every individual rat has as many observations as there are measurements. Thererore, in the long format there are 176 observations of 4 variables, which are:

- ID: unique id of participants (rats)
- Group: the treatment groups 1, 2 and 3
- Weight: rat's weight in grams
- Time: time point (days) of the weight measurement

NB: The variable 'WD' is removed from the data set as its information is already included in the 'Time' variable and thus there is no use for it. In addition, ID and Group variables are transformed to factors.

```{r message=FALSE, warning=FALSE}
# load the data in long format
rats_long <- read.table("data/rats_long.csv", sep = ";")

# prepare data for the analysis
## remove WD variable 
## Group and ID -> factors
rats_long <- rats_long %>%
  select(-WD) %>%
  mutate(Group = as.factor(Group),
         ID = as.factor(ID))

str(rats_long)
head(rats_long)
glimpse(rats_long)
```

### 6.1.2 Graphical examination of the group based differences

First, I draw a plot for every observed rat; the plot shows measured weight (y-axis) over time (x-axis) for every rat (lines). Rats in every group have gained weight during the experiment, although the growth seems to be bigger in groups 2 and 3 compared to group 1. However, the plot also illustrates that in the beginning of the experiment (time 0) the rats differ from each other in terms of the weight and bigger rats are bigger throughout the experiment. Rats in group 1 are much smaller than those in the groups 2 and 3. Also there is one much bigger individual in the group 2. 

```{r message=FALSE, warning=FALSE}

# Draw a plot for rats in each group over time
ggplot(rats_long, aes(x = Time, y = Weight, linetype = ID, col = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, times=4)) + #what this does?
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(rats_long$Weight), max(rats_long$Weight))) +
  ggtitle("Individual weight profiles over time by group for the RATS data") +
  xlab("Time (days)")

```

Next, I will standardize/center the weight variable and repeat the above-presented graphical examination. The graphical results are similar, although the trend in weight gain does not seem to be as straightforward as in the previous plot, especially regarding group 3.

```{r message=FALSE, warning=FALSE}

# Standardize the variable weight
rats_long<- rats_long %>%
  group_by(Time) %>%
  mutate(Weight_std = (Weight - mean(Weight)) / sd(Weight)) %>%
  ungroup()

# Draw the plot using the stardardixed weight variable
ggplot(rats_long, aes(x = Time, y = Weight_std, linetype = ID, col = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, times=4)) + #what this does?
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "Standardized weight") +
  ggtitle("Individual weight profiles over time by group for the RATS data after standardization") +
  xlab("Time (days)")


```

To get a better picture of the weight gain and the differences between the groups, I will next produce a graph showing average profiles for each group. Additionally I draw a graph showing side-by-side box plots of the observations at each time point. The graphs illustrate that every group has gained weight over time, although the weight gain is a bit different in different groups. The figures also show that the mean weights of groups 2 and 3 clearly differ from group 1. In addition, although there is some overlap between the groups 2 and 3 profiles it seems that they do differ from each other. Moreover, the box plots suggest the presence of multiple possible outliers at a number of time points.

```{r message=FALSE, warning=FALSE}

# Summary data with mean and standard error of weight by treatment group and time

# PROBLEM: DIFFRENT NUMBER OF RATS IN DIFFERENT GROUPS!
## how to take sqrt(n), where n = number of rats in group -> 
### n() which gives the current group size ??
rats_long_s <- rats_long %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight) /  sqrt(n())) %>%
  ungroup()

# Plot the mean profiles
ggplot(rats_long_s , aes(x = Time, y = mean, linetype = Group, shape = Group, col = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.45)) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)") +
  xlab("Time (days)") +
  ggtitle("Mean response profiles for the three groups of rats in the RATS data")

# Alternative mean profile figure: side-by-side box plots of the observations at each time point
ggplot(rats_long, aes(x = as.factor(Time), y = Weight, fill = Group)) +
  geom_boxplot(position = position_dodge(width = 0.9)) +
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position = c(0.9,0.4)) +
  scale_x_discrete(name = "Time (days") +
  ggtitle("Boxplots for the RATS data: weight over time")


```

I will continue the summary approach and look into the post-treatment values of the weight. The mean of weight (all weeks) will be the summary measure. Day 1 will be considered here as the baseline and it is excluded from the examination.

```{r}
# Create a summary data by group and subject with mean as the summary variable (ignoring baseline day 1).
rats_summary <- rats_long %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Draw a boxplot of the mean versus group
ggplot(rats_summary, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days 8-64") +
  ggtitle("Boxplots of mean summary measures for the three groups in RATS data")

```

The box plots presented above shows again that the groups differ from each other and also that every group has an outlier, although only in the case of group two the outlier is quite far from the others. Thus, I will remove the outlier of group 2 and draw the box plots again. After removing the outlier, the differences between groups are even clearer as can be seen in the picture below. 

```{r message=FALSE, warning=FALSE}
rats_summary_filt <- rats_summary %>%
  filter(mean < 550)

# Draw a boxplot of the mean versus group
ggplot(rats_summary_filt, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days 8-64") +
  ggtitle("Boxplots of mean summary measures for the three groups in RATS data \n(without outlier)")
```

### 6.1.3 Statistical examination of the group-based differences

Next, I will first fit a linear model with the mean weight as the response and group as the only explanatory variable. The results of the model are shown below.

```{r message=FALSE, warning=FALSE}

# Fit the linear model with the mean as the response and Group as explanatory
fit <- lm(mean ~ Group, data = rats_summary)
# summary of the fitted model
summary(fit)
# analysis of variance table for the fitted model 
anova(fit)

```

As can be seen from the summary table above, the positive regression coefficients of both groups are statistically significant (p < 0.001). Similarly, the analysis of variance (ANOVA) shows that the weight of the rats differ between the examined groups and that the result is highly significant (p < 0.001). Next, I will add the baseline measurement (day 1) from the original data to the summary data and include it as a covariate in the linear regression model. The results of the linear regression model and ANOVA are presented below.

```{r message=FALSE, warning=FALSE}

## load the data in wide form
rats_wide <- read.table("data/rats_wide.csv", sep = ";")

# Add the baseline (Day 1) from the original data as a new variable to the summary data
rats_summary2 <- rats_summary %>%
  mutate(baseline = rats_wide$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = rats_summary2)

# summary of the fitted model
summary(fit)
# analysis of variance table for the fitted model 
anova(fit)

```

When taking the baseline into account results are quite different! According to the above presented results from the linear regression model, the coefficient of group 2 or 3 are not statistically significant (when considering p < 0.05 as the threshold). Analysis of variance provides similar results, indicating that the differences between the groups are not statistically significant when taking the baseline into account.

## 6.2 BPRS

In this subchapter I will implement the analyses of Chapter 9 of MABS using the BPRS data.

### 6.2.1 Data

In this exercise I will use data taken from [Davis (2002)](https://link.springer.com/book/10.1007/b97287). The data is from an experimental study in which 40 males were randomly assigned to two treatment groups and each participant was rated
on a psychiatric rating scale (BPRS) measured before the treatment started (week 0) and weekly after that for eight weeks. Here, the data is already transformed from wide to long form.

```{r message=FALSE, warning=FALSE}
# load the data in long format
bprs_long <- read.table("data/bprs_long.csv", sep = ";")

# prepare data for the analysis
## remove WD variable 
## Group and ID -> factors
bprs_long <- bprs_long%>%
  select(-weeks) %>%
  mutate(treatment = as.factor(treatment),
         subject = as.factor(subject),
         id = as.factor(id))

str(bprs_long)
summary(bprs_long)
head(bprs_long)
```

BPRS data include multiple measurements from 40 subjects. As the data is transformed from a wide to long format by time (measurements), each time point forms an observation and, thus, every participant has as many observations as there are measurements. Note that the numbering of the subjects/participants goes from 1 to 20 in both treatment groups, which is why I added a new id-variable into the data which goes from 1 to 40 identifying each participants. In the long form there are 360 observations of 5 variables, which are:

- treatment: the treatment groups 1 and 2
- subject: the id of participant in given treatment group (both groups have subjects from 1 to 20)
- id: unique identifyier of participant
- bprs: results of the measurement
- week: time point (weeks) of the measurements

NB: The variable 'weeks' is removed from the data set as its information is already included in the 'week' variable and thus there is no use for it. In addition, subject and treatment variables are transformed to factors.

### 6.2.2 Graphical examination

Let's first take brief graphical examination of the data! I'll begin by plotting the BPRS-ratings over time and identifying the treatment groups in which the observations belong while ignoring the longitudinal nature of the data (i.e. which subject's observations are in question). The below-presented plot shows a downward trend in the BPRS-scores, especially regarding the first treatment group.

```{r message=FALSE, warning=FALSE}

ggplot(bprs_long, aes(x = week, y = bprs, group = id, col = treatment)) + 
  geom_text(aes(label = treatment)) + 
  scale_x_continuous(name = "Time (week)", breaks = seq(0, 11, 1)) + 
  scale_y_continuous(name = "BPRS") + 
  theme_bw() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  ggtitle("Plot of BPRS-scores over time. The treatment groups are identified by numbers and colors")

```

Next, I will take the longitudinal nature of the data into account and draw plot of individual BPRS profiles. The individual profiles show similar downward trend in the BPRS scores over time, although in the group 2 there seem to be more departures from this trend.

The next figure, that is the scatter plot matrix, demonstrates that the repeated measures of BPRS are not independent of one another, which is no surprise when considering the longitudinal nature of the data.

```{r message=FALSE, warning=FALSE}
# Plot the RATSL data
ggplot(bprs_long, aes(x = week, y = bprs, group = id, col = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 11, 1)) +
  scale_y_continuous(name = "BPRS") +
  theme_bw() + theme(legend.position = "top") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  ggtitle("Plot of individual BPRS profiles")

# load the data in wide form and draw a scatter plot matrix
bprs_wide <- read.table("data/bprs_wide.csv", sep = ";")
pairs(bprs_wide[, 3:11], cex = 0.7)
```

### 6.2.3 Regression modeling

Now I will move to a more formal analyses and fit a linear regression model including BPRS scores as the response variable and two explanatory variables, namely time and treatment group. Note: the repeated-measures structure of the data is ignored in this model. The results (see below) show that estimate of time is statistical significant (p < 0.001), meaning that the BPRS scores decrease over time. However, the regression coefficient of treatment group is not significant indicating that the BPRS scores do not differ between the groups conditional on time.

```{r message=FALSE, warning=FALSE}

# fit random intercept model 
bprs_m1 <- lm(bprs ~ week + treatment, data = bprs_long)
summary(bprs_m1)

```

The independent model conducted above assumes an independence of the measures of the response variable (bprs), which is highly unlikely here. Thus, I will fit a random intercept model, which allows the linear regression fit for each participant to differ in intercept from other participants.

```{r message=FALSE, warning=FALSE}

# access to lme4-package that provide functions for fitting mixed models
library("lme4")

# to add p-values to the summary
#install.packages("lmerTest")
library(lmerTest)

# note: in the formula the random-effects terms distinguished by vertical bars (|)
bprs_rim1 <- lmer(bprs ~ week + treatment + (1 | id), data = bprs_long, REML = FALSE)

# summary of the random intercept model
summary(bprs_rim1)



```

The results from the random intercept model are presented above. The estimates of time (i.e. week) and treatment group are quite similar to those from the previous model; the estimate of time is statistically significant, while the treatment group is not. The estimated variance of the participant (id) random effects are pretty large (97.39), which indicate considerable variation in the intercepts of the regression fits of the individual BPRS profiles. Thus, I will next fit a random intercept and random slope model, which allows the linear regression fits for each individual to differ in slope as well. This way it is possible to detect the individual differences in the participants' BPRS profiles, and also the effect of time.

Note that even though the significance of the results or coefficient of the mixed effect models is commented multiple times in the chapter 9, there are no p-values or explanations how to interpret the significance of the results, which I found very annoying. Here, I installed _lmerTest_-package which adds p-values to the lrm-summary tables.

```{r message=FALSE, warning=FALSE}

# random intercept and random slope model
bprs_rim2 <- lmer(bprs ~ week + treatment + (week | id), data = bprs_long, REML = FALSE)

# summary of the random intercept and slope model
summary(bprs_rim2)

# anova
anova(bprs_rim1, bprs_rim2)

``` 

The results from the random intercept and random slope model are presented above. The fixed effect estimate of time (i.e. week) is the same as in the random intercept model, but the fixed effect estimate of treatment group is bigger suggesting the difference between the treatment groups is larger.

The likelihood ratio test for the two models gives a chi-squared statistic of 66.66 with 2 degrees of freedom, and the associated p-value is really small (p < 0.001), meaning that the random intercept and slope model provides a better fit for the data.

Finally, I fit a random intercept and slope model that allows a interaction between treatment group and time. The results from this model are shown below. The results differ somewhat from the previous model, but more importantly the likelihood ratio test for the random intercept and slope model versus the random intercept and
slope model with interaction term shows that the latter provides worse fit for the data, and thus the results are not interpret here any further.

```{r message=FALSE, warning=FALSE}

# random intercept and random slope model with an interation term
bprs_rim3 <- lmer(bprs ~ week * treatment + (week | id), data = bprs_long, REML = FALSE)
# summary of the model
summary(bprs_rim3)
# anova
anova(bprs_rim2, bprs_rim3)

```

The very last thing to do is to draw one more figure which shows the observed and fitted BPRS profiles. Based on the above presented results I will use the random intercept and slope model without interaction term for fitted values. The two figures are presented below. They illustrate how well the model fits the observed data. Model seems to fit somewhat roughly and there are also quite big differences between the observed and fitted lines, but everyone can take a look and make a judgement of their own.

```{r message=FALSE, warning=FALSE}

# create fitted data
Fitted <- fitted(bprs_rim2)

# add the fitted values to the data frame
 bprs_long <- bprs_long %>%
   mutate(Fitted = Fitted)

graph1 <- ggplot(bprs_long , aes(x = week, y = bprs, group = id, col = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 11, 1)) +
  scale_y_continuous(name = "BPRS") +
  theme_bw() + theme(legend.position = "right") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  ggtitle("Observed BPRS")

graph2 <- ggplot(bprs_long , aes(x = week, y = Fitted, group = id, col = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 11, 1)) +
  scale_y_continuous(name = "BPRS") +
  theme_bw() + theme(legend.position = "right") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  ggtitle("Fitted BPRS")

library(ggpubr)
ggarrange(graph1, graph2,
          ncol = 2, nrow = 1, 
          common.legend = TRUE, legend = "right")

```



**THAT'S** 

*ALL* 
*FOLKS*
*AND* 
*DOCS!*


**THE END.**
